---
title: "Markov chains: Snakes and Ladders"
date: last-modified
format:
  html:
    embed-resources: true
    code-fold: true
  gfm: default
editor: visual
---

# The Rules

```{r}
#| include: false
library(tidyverse)
library(expm)
```

[Snakes and Ladders](https://en.wikipedia.org/wiki/Snakes_and_ladders) is typically played with 2 players on an $n\times n$ grid. Each user takes turns rolling a 6-sided dice, gradually progressing up the board until they reach the final square. Along the way they can land on 'snakes', which takes them down to previous squares, or 'ladders', which takes them up closer to the end.

We will look at the one player case with a few key differences:

1.  The player will start on 'square 0' outside the board, not on square 1
2.  The player can overshoot the last square (e.g., if the player is 2 away from the end, they can roll 2 or above to win)

# Representing the board

Using the board layout present in Daykin et al. (1967), the probability of going from state $i$ to state $j$ can be represented in a $101\times101$ transition matrix. While there are only 100 squares on the board, our version requires an initial 'zero' state.

Let $p_{ij}$ be the probability of going from state $i$ to state $j$. Then:
$$P=
\begin{pmatrix}
p_{00} & p_{01} & p_{02} & \cdots & p_{0\ 100}\\
p_{10} & p_{11} & p_{12} & \cdots & p_{1\ 100}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
p_{100\ 0} & p_{100\ 1} & p_{100\ 2} & \cdots & p_{100\ 100}\\
\end{pmatrix}$$

Below are the first and last 5 rows of our matrix.

```{r, board generation}
# size of board
n = 100

# number of sides of dice
d = 6

# first number is the 'entry' square, second number is 'exit' square
ladders = list(c(6,23),c(8,30),c(13,47),c(20,39),c(33,70),
               c(37,75),c(41,62),c(57,83),c(66,89),c(77,96))

snakes = list(c(27,10),c(55,16),c(61,14),c(69,50),c(79,5),
              c(81,44),c(87,31),c(91,25),c(95,49),c(97,59))

board = matrix(0,n,n)

# create simple board with no ladders
for (i in 1:n) {
  for (j in 0:min(n-i,d-1)) {
    board[i,j+i] = 1/d 
  } 
}

# add initial and absorbing state
board = board %>%  
  cbind(rep(0,n), .) %>% 
  rbind(c(rep(0,n),1))

# fill board with probabilities
board[,n+1][max(1,(n-(d-1))):n] = (max(1,1+d-n):d)/d

rownames(board) = 0:(n)
colnames(board) = 0:(n)

# vector of states to be deleted
delete = c()

# add in ladders
for (i in ladders) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

# add in snakes
for (i in snakes) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

board = board[-(delete+1),-(delete+1)]
knitr::kable(head(board), digits=3)
knitr::kable(tail(board), digits=3)
```

Note that some states are missing. This is because each ladder/snake reduces the number of possible states by one. Consider the snake beginning from square 97 to square 59. Landing on square 97 is equivalent to landing on square 59. Thus, the number of states is reduced by one. This holds true for ladders.

Since there are a total of 20 snakes and ladders, this leaves us with a final $81\times81$ transition matrix. 

```{r}

# simulate discrete Markov chains according to transition matrix P
simulatemc = function(tm) {
  
  # vector of states over time t
  states = c()

  # initialise variable for first state 
  states[1] = 1
  
  t = 1
  while (states[t] != nrow(tm)) {
    t = t + 1
    # probability vector to simulate next state X_{t+1}
    p = tm[states[t-1], ]
    
    # draw from multinomial and determine state
    states[t] <-  which(rmultinom(1, 1, p) == 1)
  }
  return(states)
}

```

The `echo: false` option disables the printing of code (only output is displayed).

```{r}

avg = c()
#lln = c()
for (j in 1:100) {
  run_time = c()
  for (i in 1:100) {
    run_time = c(run_time, length(simulatemc(board))-1)
    #lln = c(lln, mean(run_time))
  }
  avg = c(avg, mean(run_time))
}

difference = c()
for (i in snakes) {
  difference = c(difference, i[1]-i[2])
}
difference

difference = c()
for (i in ladders) {
  difference = c(difference, i[2]-i[1])
}
difference

qqnorm(avg) 
qqline(avg)
t.test(avg, mu=47.98, alternative="two.sided")
mean(run_time)
sd(run_time)

player1 = simulatemc(board)
player2 = simulatemc(board)
plot(player1, type='o', col='red', ylab="Position on board", xlab="Turns")
lines(player2, type='o', col='blue')
player1
player2

#plot(lln, type='l', ylim = c(45,55))
#hist(run_time, breaks=50)

plot(seq(20, 80, length=1000), dnorm(seq(20, 80, length=1000), mean=mean(run_time), sd=sd(run_time)), type="l", lwd=1)
```

```{r}
# expected number of visits from transient state i to transient state j
fundamental = solve(diag(nrow(board)-1) - board[1:(nrow(board)-1),1:(nrow(board)-1)])

ex_steps = fundamental %*% matrix(1,nrow(board)-1,1) # expected number of steps to absorbing state from state i
fundamental[1,] #expected number of visits to state i from the starting position

(fundamental-diag(nrow(board)-1)) %*% solve(diag(diag(fundamental),nrow(board)-1,nrow(board)-1))

plot(ex_steps, type="l")

# test = data.frame(visits=fundamental[1,],row=paste("Row",1:10),col=paste("Col",rep(1:8,each=10)),pos=rownames(fundamental))

test = data.frame(visits=fundamental[1,], pos=as.numeric(rownames(fundamental))) %>% 
  complete(pos = seq(min(pos), max(pos)))

for (i in ladders) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

for (i in snakes) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

test$row = rep(1:10,10)
test$col = rep(1:10,each=10)

ggplot(test, aes(x=row,y=col,fill=visits)) +
  geom_tile() +
  geom_text(aes(label=round(visits,3)), color="white")

plot(test$visits,type="p")

for (i in ladders) {
  abline(v=i[1]+1, lty="dotted")
}
for (i in snakes) {
  abline(v=i[1]+1, lty="dotted")
}

```

```{r}
states = 1:80

#board[1:(nrow(board)-1),nrow(board)]

init = c(1,rep(0,nrow(board)-2))
prob = c()

for (i in 1:(500)) {
  num = matrix(init, nrow =1, ncol=nrow(board)-1) %*% (board[1:(nrow(board)-1),1:(nrow(board)-1)] %^% (i-1)) %*% board[1:(nrow(board)-1),nrow(board)]
  prob = c(prob, as.numeric(num))
}

plot(prob, type="l")
abline(v=48)
sum(prob)

#sum((1:length(prob))*prob)
```

# References

1.  Daykin, D. E., Jeacocke, J. E., & Neal, D. G. (1967). Markov Chains And Snakes And Ladders. The Mathematical Gazette, 51(378), 313–317. <https://doi.org/10.2307/3612943>

# Appendix

## Appendix A: Ladder and Snake positions

Note: The first number indicates the entry point and the second number the exit point

10 snakes: (27,10), (55,16), (61,14), (69,50), (79,5), (81,44), (87,31), (91,25), (95,49), (97,59)

10 ladders: (6,23), (8,30), (13,47), (20,39), (33,70), (37,75), (41,62), (57,83), (66,89), (77,96)
