---
title: "Markov chains: Snakes and Ladders"
format:
  html:
    embed-resources: true
    code-fold: true
  gfm: default
editor: visual
---

## Introduction

```{r}
#| include: false
library(tidyverse)
library(expm)
```

The classic Snakes and Ladders board game can be represented by an $n$ by $n$ Markov Chain

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
# size of board
n = 100

# number of sides of dice
d = 6

ladders = list(c(6,23),c(8,30),c(13,47),c(20,39),c(33,70),
               c(37,75),c(41,62),c(57,83),c(66,89),c(77,96))

snakes = list(c(27,10),c(55,16),c(61,14),c(69,50),c(79,5),
              c(81,44),c(87,31),c(91,25),c(95,49),c(97,59))

#ladders = list()
#snakes = list()

board = matrix(0,n,n)

# create simple board with no ladders
for (i in 1:n) {
  for (j in 0:min(n-i,d-1)) {
    board[i,j+i] = 1/d 
  } 
}

board = board %>%  
  cbind(rep(0,n), .) %>% 
  rbind(c(rep(0,n),1))

#if n < d: start from 1+(d-n)
#else start from 1
board[,n+1][max(1,(n-(d-1))):n] = (max(1,1+d-n):d)/d

rownames(board) = 0:(n)
colnames(board) = 0:(n)

delete = c()

# add in ladders
for (i in ladders) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

# add in snakes
for (i in snakes) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

delete = delete+1

board = board[-(delete),-(delete)]
```

You can add options to executable code like this

```{r}

# simulate discrete Markov chains according to transition matrix P
simulatemc = function(tm) {
  
  # vector of states over time t
  states = c()

  # initialise variable for first state 
  states[1] = 1
  
  t = 1
  while (states[t] != nrow(tm)) {
    t = t + 1
    # probability vector to simulate next state X_{t+1}
    p = tm[states[t-1], ]
    
    # draw from multinomial and determine state
    states[t] <-  which(rmultinom(1, 1, p) == 1)
  }
  return(states)
}

```

The `echo: false` option disables the printing of code (only output is displayed).

```{r}

avg = c()
#lln = c()
for (j in 1:100) {
  run_time = c()
  for (i in 1:100) {
    run_time = c(run_time, length(simulatemc(board))-1)
    #lln = c(lln, mean(run_time))
  }
  avg = c(avg, mean(run_time))
}

difference = c()
for (i in snakes) {
  difference = c(difference, i[1]-i[2])
}
difference

difference = c()
for (i in ladders) {
  difference = c(difference, i[2]-i[1])
}
difference

qqnorm(avg) 
qqline(avg)
t.test(avg, mu=47.98, alternative="two.sided")
mean(run_time)
sd(run_time)

player1 = simulatemc(board)
player2 = simulatemc(board)
plot(player1, type='o', col='red', ylab="Position on board", xlab="Turns")
lines(player2, type='o', col='blue')
player1
player2

#plot(lln, type='l', ylim = c(45,55))
#hist(run_time, breaks=50)

plot(seq(20, 80, length=1000), dnorm(seq(20, 80, length=1000), mean=mean(run_time), sd=sd(run_time)), type="l", lwd=1)
```

```{r}
# expected number of visits from transient state i to transient state j
fundamental = solve(diag(nrow(board)-1) - board[1:(nrow(board)-1),1:(nrow(board)-1)])

ex_steps = fundamental %*% matrix(1,nrow(board)-1,1) # expected number of steps to absorbing state from state i
fundamental[1,] #expected number of visits to state i from the starting position

(fundamental-diag(nrow(board)-1)) %*% solve(diag(diag(fundamental),nrow(board)-1,nrow(board)-1))

plot(ex_steps, type="l")

# test = data.frame(visits=fundamental[1,],row=paste("Row",1:10),col=paste("Col",rep(1:8,each=10)),pos=rownames(fundamental))

test = data.frame(visits=fundamental[1,], pos=as.numeric(rownames(fundamental))) %>% 
  complete(pos = seq(min(pos), max(pos)))

for (i in ladders) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

for (i in snakes) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

test$row = rep(1:10,10)
test$col = rep(1:10,each=10)

ggplot(test, aes(x=row,y=col,fill=visits)) +
  geom_tile() +
  geom_text(aes(label=round(visits,3)), color="white")

plot(test$visits,type="p")

for (i in ladders) {
  abline(v=i[1]+1, lty="dotted")
}
for (i in snakes) {
  abline(v=i[1]+1, lty="dotted")
}

```

```{r}
states = 1:80

#board[1:(nrow(board)-1),nrow(board)]

init = c(1,rep(0,nrow(board)-2))
prob = c()

for (i in 1:(500)) {
  num = matrix(init, nrow =1, ncol=nrow(board)-1) %*% (board[1:(nrow(board)-1),1:(nrow(board)-1)] %^% (i-1)) %*% board[1:(nrow(board)-1),nrow(board)]
  prob = c(prob, as.numeric(num))
}

plot(prob, type="l")
abline(v=48)
sum(prob)

#sum((1:length(prob))*prob)
```
