---
title: "Markov chains: Snakes and Ladders"
date: last-modified
format:
  html:
    embed-resources: true
    code-fold: true
  gfm: default
editor: visual
---

# The Rules

```{r, libraries}
#| include: false
options(scipen = 10)
library(tidyverse)
library(expm)
```

[Snakes and Ladders](https://en.wikipedia.org/wiki/Snakes_and_ladders) is typically played with 2 players on an $n\times n$ grid. Each user takes turns rolling a 6-sided dice, gradually progressing up the board until they reach the final square. Along the way they can land on 'snakes', which takes them down to previous squares, or 'ladders', which takes them up closer to the end.

We will look at the one player case with a few key differences:

1.  The player will start on 'square 0' outside the board, not on square 1
2.  The player can overshoot the last square (e.g., if the player is 2 away from the end, they can roll 2 or above to win)

# Representing the board

Using the board layout present in Daykin et al. (1967), the probability of going from state $i$ to state $j$ can be represented in a $101\times101$ transition matrix. While there are only 100 squares on the board, our version requires an initial 'zero' state.

Let $p_{ij}$ be the probability of going from state $i$ to state $j$. Then: $$P=
\begin{pmatrix}
p_{00} & p_{01} & p_{02} & \cdots & p_{0\ 100}\\
p_{10} & p_{11} & p_{12} & \cdots & p_{1\ 100}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
p_{100\ 0} & p_{100\ 1} & p_{100\ 2} & \cdots & p_{100\ 100}\\
\end{pmatrix}$$

Below are the first and last 5 rows of our matrix.

```{r, board generation}
# size of board
n = 100

# number of sides of dice
d = 6

# first number is the 'entry' square, second number is 'exit' square
ladders = list(c(6,23),c(8,30),c(13,47),c(20,39),c(33,70),
               c(37,75),c(41,62),c(57,83),c(66,89),c(77,96))

snakes = list(c(27,10),c(55,16),c(61,14),c(69,50),c(79,5),
              c(81,44),c(87,31),c(91,25),c(95,49),c(97,59))

board = matrix(0,n,n)

# create simple board with no ladders
for (i in 1:n) {
  for (j in 0:min(n-i,d-1)) {
    board[i,j+i] = 1/d 
  } 
}

# add initial and absorbing state
board = board %>%  
  cbind(rep(0,n), .) %>% 
  rbind(c(rep(0,n),1))

# fill board with probabilities
board[,n+1][max(1,(n-(d-1))):n] = (max(1,1+d-n):d)/d

rownames(board) = 0:(n)
colnames(board) = 0:(n)

# vector of states to be deleted
delete = c()

# add in ladders
for (i in ladders) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

# add in snakes
for (i in snakes) {
  for (j in (i[1]-(d-1)):i[1]) {
    board[j,][i[2]+1] = board[j,][i[2]+1] + 1/d
  }
  delete = c(delete, i[1])
}

board = board[-(delete+1),-(delete+1)]
knitr::kable(head(board), digits=3)
knitr::kable(tail(board), digits=3)
```

Note that some states are missing. This is because each ladder/snake reduces the number of possible states by one.

Consider the snake beginning from square 97 to square 59. Landing on square 97 is equivalent to landing on square 59. Thus, the number of states is reduced by one. This holds true for ladders.

Since there are a total of 20 snakes and ladders, this leaves us with a final $81\times81$ transition matrix.

# Expected number of turns to finish

Since our chain constitutes an absorbing Markov chain, we have the canonical form of a transition matrix:
$$P=\begin{bmatrix}
Q & R\\
0 & I\\
\end{bmatrix}$$
From this, we can get the fundamental matrix, $N$ given by:
$$N=(I-Q)^{-1}$$
This represents the number of expected visits from transient state $i$ to transient state $j$. To get the expected absorbing time (or time to finish),$t$ , we multiply the fundamental matrix by a column vector of 1s like so $t=N1$

```{r, expected number of turns}
fundamental = solve(diag(nrow(board)-1) - board[1:(nrow(board)-1),1:(nrow(board)-1)])
ex_steps = fundamental %*% matrix(1,nrow(board)-1,1)
plot(ex_steps, xlab="State", ylab="Expected steps to win", main="Number of turns to win based on state")
head(ex_steps)
```

Therefore, on average, it would take around `{r}round(ex_steps[1],2)` turns to win. If we were to start on square 1, it would take `{r}round(ex_steps[2],2)` turns to win and so on.

# Simulating games

Let us simulate a few games of Snakes and Ladders. I've highlighted the shortest game blue and the longest game red.

```{r, simulation}
# simulate discrete Markov chains according to transition matrix P
simulate.mc = function(tm) {
  
  # vector of states over time t
  states = c()

  # initialise variable for first state 
  states[1] = 1
  
  t = 1
  while (states[t] != nrow(tm)) {
    t = t + 1
    # probability vector to simulate next state X_{t+1}
    p = tm[states[t-1], ]
    
    # draw from multinomial and determine state
    states[t] <-  which(rmultinom(1, 1, p) == 1)
  }
  return(states)
}

games = list()
run_time = c()

# for reproducibility
set.seed(1984)

for (i in 1:20) {
  current_game = simulate.mc(board)
  
  # remove 1 to account for starting turn
  run_time = c(run_time, length(current_game)-1)
  games[[i]] = current_game
}

longest = which.max(run_time)
shortest = which.min(run_time)

plot(games[[longest]], type="l", col="red", xlab="Turns", ylab="State", main="20 simulated runs of Snakes and Ladders")
lines(games[[shortest]], col="blue")
for (i in 2:length(games)) {
  if (!(i == longest | i == shortest)) {
    lines(games[[i]], col=alpha("black",0.1))
  }
}
paste("Longest game:", run_time[longest], "turns")
paste("Shortest game:", run_time[shortest], "turns")
paste("Average game:", mean(run_time), "turns")
```

Will the simulated average converge to our calculated theoretical value? According to the Law of Large numbers, repeating the simulated trials over a large number of samples should cause our mean to converge to the true value (marked by the red dotted line).

```{r, t.test}
#| cache: true

trials = 100000

run_time = numeric(trials)
avg = numeric(trials)

for (i in 1:trials) {
  current_game = simulate.mc(board)
    
  # remove 1 to account for starting turn
  run_time[i] = length(current_game)-1
  avg[i] = mean(run_time[1:i])
}

plot(avg, type="l", ylim=c(47,49), xlim=c(5000,100000), xlab="Number of trials", ylab="Mean steps to finish", main="Average time to finish by number of trials")
abline(h=ex_steps[1], lty=2, col="red")

paste("Longest game:", max(run_time), "turns")
paste("Average game:", mean(run_time), "turns")
```


```{r}
# expected number of visits from transient state i to transient state j
fundamental = solve(diag(nrow(board)-1) - board[1:(nrow(board)-1),1:(nrow(board)-1)])

ex_steps = fundamental %*% matrix(1,nrow(board)-1,1) # expected number of steps to absorbing state from state i
fundamental[1,] #expected number of visits to state i from the starting position

(fundamental-diag(nrow(board)-1)) %*% solve(diag(diag(fundamental),nrow(board)-1,nrow(board)-1))

# test = data.frame(visits=fundamental[1,],row=paste("Row",1:10),col=paste("Col",rep(1:8,each=10)),pos=rownames(fundamental))

test = data.frame(visits=fundamental[1,], pos=as.numeric(rownames(fundamental))) %>% 
  complete(pos = seq(min(pos), max(pos)))

for (i in ladders) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

for (i in snakes) {
  test[test$pos==i[1],]$visits = test[test$pos==i[2],]$visits
}

test$row = rep(1:10,10)
test$col = rep(1:10,each=10)

ggplot(test, aes(x=row,y=col,fill=visits)) +
  geom_tile() +
  geom_text(aes(label=round(visits,3)), color="white")

plot(test$visits,type="p")

for (i in ladders) {
  abline(v=i[1]+1, lty="dotted")
}
for (i in snakes) {
  abline(v=i[1]+1, lty="dotted")
}

```

```{r}
states = 1:80

#board[1:(nrow(board)-1),nrow(board)]

init = c(1,rep(0,nrow(board)-2))
prob = c()

for (i in 1:(500)) {
  num = matrix(init, nrow =1, ncol=nrow(board)-1) %*% (board[1:(nrow(board)-1),1:(nrow(board)-1)] %^% (i-1)) %*% board[1:(nrow(board)-1),nrow(board)]
  prob = c(prob, as.numeric(num))
}

plot(prob, type="l")
abline(v=48)
sum(prob)

#sum((1:length(prob))*prob)
```

# References

1.  Daykin, D. E., Jeacocke, J. E., & Neal, D. G. (1967). Markov Chains And Snakes And Ladders. The Mathematical Gazette, 51(378), 313–317. <https://doi.org/10.2307/3612943>

# Appendix

## Appendix A: Ladder and Snake positions

Note: The first number indicates the entry point and the second number the exit point

10 snakes: (27,10), (55,16), (61,14), (69,50), (79,5), (81,44), (87,31), (91,25), (95,49), (97,59)

10 ladders: (6,23), (8,30), (13,47), (20,39), (33,70), (37,75), (41,62), (57,83), (66,89), (77,96)

```{r}
#| eval: false
#| include: false

# difference = c()
# for (i in snakes) {
#   difference = c(difference, i[1]-i[2])
# }
# difference
# 
# difference = c()
# for (i in ladders) {
#   difference = c(difference, i[2]-i[1])
# }
# difference
# 
# player1 = simulate.mc(board)
# player2 = simulate.mc(board)
# plot(player1, type='o', col='red', ylab="Position on board", xlab="Turns")
# lines(player2, type='o', col='blue')
# player1
# player2
```

